---
title: "04-lab-Data-Visualization"
author: "Flemming Wu"
date: "2022-09-14"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```


### 1. Read in the data

```{r, message = FALSE}
if(!file.exists("met_all.gz")){
  download.file(
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
    destfile = "met_all.gz",
    method = "libcurl",
    timeout = 60
  )
}

met <- data.table::fread("met_all.gz")
```
```{r, message = FALSE}
library(tidyverse)
library(data.table)
```

### 2. Preparing the data

Remove temperatures less than -17C.\
Make sure there are no missing data in the key variables coded as 9999, 999, etc.\
Generate a date variable using the functions as.Date() (hint: You will need the following to create a date paste(year, month, day, sep = "-")).\
Using the data.table::week function, keep the observations of the first week of the month.\
Compute the mean by station of the variables temp, rh, wind.sp, vis.dist, dew.point, lat, lon, and elev.\
Create a region variable for NW, SW, NE, SE based on lon = -98.00 and lat = 39.71 degrees.\
Create a categorical variable for elevation as in the lecture slides.


```{r Checking dimensions}
dim(met)
```
```{r Checking key vars}
summary(met$elev)
```

```{r}
summary(met$temp)
```

```{r Remove temperatures less than -17C and make sure there are no missing data in the key variables coded as 9999, 999, etc}
met <- met[met$temp > -17][elev == 9999.0, elev := NA]
```

```{r Generate a date variable}
met$date = paste(met$year, met$month, met$day, sep = "-")
met <- mutate(met, date = as.Date(date, "%Y-%m-%d"))
```


```{r}
met$week <- data.table::week(met$date)
```

```{r Keep the observations of the first week of the month}
met <- met[met$week == min(met$week)]
```

```{r Compute the mean by station of the variables temp, rh, wind.sp, vis.dist, dew.point, lat, lon, and elev.}

met_avg <- met[,.(
  temp     = mean(temp,na.rm=TRUE),
  rh       = mean(rh,na.rm=TRUE),
  wind.sp  = mean(wind.sp,na.rm=TRUE),
  vis.dist = mean(vis.dist,na.rm=TRUE),
  dew.point = mean(dew.point,na.rm=TRUE),
  lat      = mean(lat),
  lon      = mean(lon), 
  elev     = mean(elev,na.rm=TRUE)
), by="USAFID"]

```

```{r Create a region variable for NW, SW, NE, SE}

met_avg[, region := fifelse(lon > -98 & lat > 39.71, "NE", 
                fifelse(lon < -98 & lat > 39.71, "NW",
                fifelse(lon < -98 & lat <= 39.71, "SW", "SE")))
                ]

table(met_avg$region)

```

```{r Create a categorical variable for elevation as in the lecture slides}
met_avg[, elev_cat := fifelse(elev > 252, "high", "low")]
```



### 3. Use geom_violin() to examine the wind speed and dew point temperature by region
Use facets.\
Make sure to deal with NA category.\
Describe what you observe in the graph.\

```{r Dew Point by region}
met_avg[!is.na(region)] %>%
  ggplot() +
  geom_violin(mapping = aes(x = 1, y = dew.point, color = region, fill = region)) +
  facet_wrap(~ region, nrow = 1)
```
The highest dew point temperature is recorded in the southeast.

```{r Wind speed by region}
met_avg[!is.na(region) & !is.na(wind.sp)] %>%
  ggplot() +
  geom_violin(aes(x =  1, y = wind.sp, color = region, fill = region)) +
  facet_wrap(~ region, nrow = 1)
```
The highest wind speed was recorded in the northeast.


### 4. Use geom_jitter with geom_smooth to examine the association between dew point temperature and wind speed by region
Colour points by region.\
Make sure to deal with NA category.\
Fit a linear regression line by region.\
Describe what you observe in the graph.\


```{r}
met_avg[!is.na(region) & !is.na(wind.sp)] %>%
  ggplot(mapping = aes(x = wind.sp, y = dew.point, color = region)) +
  geom_point(mapping = aes(color = region)) +
  geom_smooth(method = lm, mapping = aes(linetype = region)) +
  facet_wrap(~ region, nrow = 2)
```
It appears that the southwest region is the only region that sees a decrease in dewpoint with increasing wind speed.


### 5. Use geom_bar to create barplots of the weather stations by elevation category coloured by region

Bars by elevation category using position="dodge".\
Change colours from the default. Colour by region using scale_fill_brewer see this.\
Create nice labels on axes and add a title.\
Describe what you observe in the graph.\
Make sure to deal with NA.

```{r}
met_avg[!is.na(elev_cat)] %>%
  ggplot() +
  geom_bar(mapping = aes(x = elev_cat, col = region, fill = region), position = "dodge") +
  scale_fill_brewer(palette ="Set2") +
  xlab("Elevation Category") +
  ylab("Number of Stations") +
  ggtitle("Number of Stations by Elevation Category")
  
```
The majority of stations in the southeast are at a low elevation, whereas in the northwest and southwest more stations are at a higher elevation. For the northeast, the stations are about evenly distributed between high and low elevations.


### 6. Use stat_summary to examine mean dew point and wind speed by region with standard deviation error bars

Make sure to remove NA.\
Use fun.data=“mean_sdl” in stat_summary.\
Add another layer of stats_summary but change the geom to "errorbar" (see the help).\
Describe the graph and what you observe.\
Dew point temperature is…\
Wind speed is…


```{r}
met_avg[!is.na(dew.point)] %>%
  ggplot(mapping = aes(x=region, y=dew.point)) +
  stat_summary(fun.data = "mean_sdl", geom = "errorbar")
```
The west seems to have lower dew points on average. Additionally, the range of error in the west is much higher than in the east, where dew point readings seem to lie close to the mean.


### 7. Make a map showing the spatial trend in relative h in the US

Make sure to remove NA/\
Use leaflet().\
Make a colour palette with custom colours.\
Use addMarkers to include the top 10 places in relative h (hint: this will be useful rank(-rh) <= 10).\
Add a legend.\
Describe trend in RH across the US.

```{r Filter top 10 places in the US by rh}
top10rh <- met_avg[rank(-rh) <= 10]

#top10rh
```


```{r, message = FALSE}
library(leaflet)
library(webshot)

rh.pal <- colorNumeric(c('darkgreen','goldenrod','brown'), domain=met_avg$rh)
#rh.pal
```

```{r}
rhmap <- leaflet(met_avg) %>% 
  # The looks of the Map
  addProviderTiles('CartoDB.Positron') %>% 
  # Some circles
  addCircles(
    lat = ~lat, lng=~lon,
    label = ~paste0(rh), color = ~ rh.pal(rh),
    opacity = 1, fillOpacity = 1, radius = 500
    ) %>%
  # And a pretty legend
  addLegend('bottomleft', pal=rh.pal, values=met_avg$rh,
          title='Relative Humidity', opacity=1)

rhmap
```

### 8. Use a ggplot extension

Pick and extension (except cowplot) from here and make a plot of your choice using the met data (or met_avg).\
Might want to try examples that come with the extension first (e.g. ggtech, gganimate, ggforce).

```{r}
library(ggbeeswarm)

met_avg[!is.na(dew.point)] %>%
  ggplot(aes(x = region, y = dew.point, color = elev_cat)) +
  geom_beeswarm(size = 0.5) +
  ggtitle("Beeswarm Plot")
```
The beeswarm plot is a hybrid between a violin plot and scatter plot. It helps visualize the density of the data at each point (similar to a violin plot), while still showing each data point individually.
From this plot, it can be observed that dew points are higher for low elevations in the southeast, while the converse is true for the northeast. The southwest has about an equal distribution of dew points in both its elevation categories, and the northwest lacks data at low elevations.