---
title: "04-lab-Data-Visualization"
author: "Flemming Wu"
date: "2022-09-14"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### 1. Read in the data

```{r, message = FALSE}
if(!file.exists("met_all.gz")){
  download.file(
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
    destfile = "met_all.gz",
    method = "libcurl",
    timeout = 60
  )
}

met <- data.table::fread("met_all.gz")
```
```{r, message = FALSE}
library(tidyverse)
library(data.table)
```

### 2. Preparing the data

```{r Checking dimensions}
dim(met)
```
```{r Checking key vars}
summary(met$elev)
```

```{r}
summary(met$temp)
```

```{r Remove temperatures less than -17C and make sure there are no missing data in the key variables coded as 9999, 999, etc}
met <- met[met$temp > -17][elev == 9999.0, elev := NA]
```

```{r Generate a date variable}
met$date = paste(met$year, met$month, met$day, sep = "-")
met <- mutate(met, date = as.Date(date, "%Y-%m-%d"))
```


```{r}
met$week <- data.table::week(met$date)
```

```{r Keep the observations of the first week of the month}
met <- met[met$week == min(met$week)]
```

```{r Compute the mean by station of the variables temp, rh, wind.sp, vis.dist, dew.point, lat, lon, and elev.}

met_avg <- met[,.(
  temp     = mean(temp,na.rm=TRUE),
  rh       = mean(rh,na.rm=TRUE),
  wind.sp  = mean(wind.sp,na.rm=TRUE),
  vis.dist = mean(vis.dist,na.rm=TRUE),
  dew.point = mean(dew.point,na.rm=TRUE),
  lat      = mean(lat),
  lon      = mean(lon), 
  elev     = mean(elev,na.rm=TRUE)
), by="USAFID"]

```

```{r Create a region variable for NW, SW, NE, SE}

met_avg[, region := fifelse(lon > -98 & lat > 39.71, "NE", 
                fifelse(lon < -98 & lat > 39.71, "NW",
                fifelse(lon < -98 & lat <= 39.71, "SW", "SE")))
                ]

table(met_avg$region)

```

```{r Create a categorical variable for elevation as in the lecture slides}
met_avg[, elev_cat := fifelse(elev > 252, "high", "low")]
```



### 3. Use geom_violin() to examine the wind speed and dew point temperature by region
Use facets
Make sure to deal with NA category
Describe what you observe in the graph

```{r Dew Point by region}
met_avg[!is.na(region)] %>%
  ggplot() +
  geom_violin(mapping = aes(x = 1, y = dew.point, color = region, fill = region)) +
  facet_wrap(~ region, nrow = 1)
```
The highest dew point temperature is recorded in the southeast.

```{r Wind speed by region}
met_avg[!is.na(region) & !is.na(wind.sp)] %>%
  ggplot() +
  geom_violin(aes(x =  1, y = wind.sp, color = region, fill = region)) +
  facet_wrap(~ region, nrow = 1)
```
The highest wind speed was recorded in the northeast.


### 4. Use geom_jitter with geom_smooth to examine the association between dew point temperature and wind speed by region
Colour points by region
Make sure to deal with NA category
Fit a linear regression line by region
Describe what you observe in the graph


```{r}
met_avg[!is.na(region) & !is.na(wind.sp)] %>%
  ggplot(mapping = aes(x = wind.sp, y = dew.point, color = region)) +
  geom_point(mapping = aes(color = region)) +
  geom_smooth(method = lm, mapping = aes(linetype = region)) +
  facet_wrap(~ region, nrow = 2)
```
It appears that the southwest region is the only region that sees a decrease in dewpoint with increasing wind speed.


### 5. Use geom_bar to create barplots of the weather stations by elevation category coloured by region

Bars by elevation category using position="dodge"
Change colours from the default. Colour by region using scale_fill_brewer see this
Create nice labels on axes and add a title
Describe what you observe in the graph
Make sure to deal with NA

```{r}
met_avg[!is.na(elev_cat)] %>%
  ggplot() +
  geom_bar(mapping = aes(x = elev_cat, col = region))
```

### 6. Use stat_summary to examine mean dew point and wind speed by region with standard deviation error bars

Make sure to remove NA
Use fun.data=“mean_sdl” in stat_summary
Add another layer of stats_summary but change the geom to "errorbar" (see the help).
Describe the graph and what you observe
Dew point temperature is…
Wind speed is…


```{r}
met_avg[!is.na(dew.point)] %>%
  ggplot(mapping = aes(x=region, y=dew.point)) +
  stat_summary(fun.data = "mean_sdl", geom = "errorbar")
```
The west seems to have lower dew points on average


### 7. Make a map showing the spatial trend in relative h in the US

Make sure to remove NA
Use leaflet()
Make a colour palette with custom colours
Use addMarkers to include the top 10 places in relative h (hint: this will be useful rank(-rh) <= 10)
Add a legend
Describe trend in RH across the US

```{r}
top10rh <- met_avg[rank(-rh) <= 10]

top10rh
```


```{r}
library(leaflet)
library(webshot)


rh.pal <- colorNumeric(c('darkgreen','goldenrod','brown'), domain=met_avg$rh)
rh.pal


```

```{r}
rhmap <- leaflet(met_avg) %>% 
  # The looks of the Map
  addProviderTiles('CartoDB.Positron') %>% 
  # Some circles
  addCircles(
    lat = ~lat, lng=~lon,
    label = ~paste0(rh), color = ~ rh.pal(rh),
    opacity = 1, fillOpacity = 1, radius = 500
    ) %>%
  # And a pretty legend
  addLegend('bottomleft', pal=rh.pal, values=met_avg$rh,
          title='Relative Humidity', opacity=1)

rhmap
```

