---
title: "Lab 01: Hello R"
author: "Flemming Wu"
date: "8/24/2022"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    fig_width: 3
    fig_height: 3
    highlight: tango
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-libraries}
#need to load the libraries in the md file (not console) for knitr to work
library(tidyverse)
library(datasauRus)
```


## Exercise 1

How many rows and columns are in the dataset datasaurus_dozen?

```{r}
dim(datasaurus_dozen)
```
1846 rows, and 3 columns

What are the variables included in the data frame?

```{r dataset-object}
datasaurus_dozen
```
name of dataset (character), x (double), y (double)

```{r}
datasaurus_dozen %>%
  count(dataset)
```
142 entries for each of the datasets


## Exercise 2

Plot y vs x for the dino dataset. Then, calculate the correlation coefficient between x and y for this dataset.

```{r run-dino}
dino_data <- datasaurus_dozen %>%
  filter(dataset == "dino")

ggplot(data = dino_data, mapping = aes(x = x, y = y)) +
  geom_point()
```

```{r}
dino_data %>%
  summarize(r = cor(x, y))
```


## Exercise 3

Plot y vs x for the star dataset.


```{r run-star}
star_data <- datasaurus_dozen %>%
  filter(dataset == "star")

ggplot(data = star_data, mapping = aes(x = x, y = y)) +
  geom_point()
```

```{r}
star_data %>%
  summarize(r = cor(x, y))
```
Very similar correlation coefficient with the dino dataset


## Exercise 4

Plot y vs x for the circle dataset.

```{r run-circle}
circle_data <- datasaurus_dozen %>%
  filter(dataset == "circle")

ggplot(data = circle_data, mapping = aes(x = x, y = y)) +
  geom_point()
```

```{r}
circle_data %>%
  summarize(r = cor(x, y))
```
Also has a similar correlation coefficient to those of the star and dino datasets

## Exercise 5

Put all dataset figures in 1 plot
```{r run-all, fig.width=6, fig.height=6}
ggplot(datasaurus_dozen, aes(x = x, y = y, color = dataset))+
  geom_point()+
  facet_wrap(~ dataset, ncol = 3) +
  theme(legend.position = "none")
```

```{r}
datasaurus_dozen %>%
  group_by(dataset) %>%
  summarize(r = cor(x, y))
```

```{r}
sessionInfo()
```

