---
title: "Assignment 02: Data Viz and Wrangling"
author: "Flemming Wu"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true
---

For this assignment, we will be analyzing data from USC’s Children’s Health Study. The learning objectives are to conduct data wrangling and visualize the data with key questions in mind.

### Data Wrangling
You will need to download two datasets from https://github.com/USCbiostats/data-science-data. The individual and regional CHS datasets in 01_chs. The individual data includes personal and health characteristics of children in 12 communities across Southern California. The regional data include air quality measurements at the community level. Once downloaded, you can merge these datasets using the location variable. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
setwd("~/Desktop/PM566/GitHub/02-assignment")
```

Load libraries
```{r load libraries, message = FALSE}
library(data.table)
library(tidyverse)
library(dtplyr)
library(leaflet)
```

Download data
```{r download data}
if(!file.exists("individual.csv")){
  download.file(url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/01_chs/chs_individual.csv",
               destfile = "individual.csv",
               method = "libcurl",
               timeout = 60)
}

if(!file.exists("regional.csv")){
  download.file(url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/01_chs/chs_regional.csv",
               destfile = "regional.csv",
               method = "libcurl",
               timeout = 60)
}

#Read data tables
individual <- fread("individual.csv")
regional <- fread("regional.csv")

```

######  1. After merging the data, make sure you don’t have any duplicates by counting the number of rows. Make sure it matches.\
In the case of missing values, impute data using the average within the variables “male” and “hispanic.” If you are interested (and feel adventurous) in the theme of Data Imputation, take a look at this paper on “Multiple Imputation” using the Amelia R package here.\
  
Merge data tables and ensure dimensions are correct.
```{r merging data}
#Check rows
dim(individual) #merged data table should contain 1200 rows
dim(regional) #merged data table should have 23 + 27 - 1 columns

dat <- merge(x = individual, y = regional, by.x = "townname", by.y = "townname")

#Check dimensions
dim(dat) #1200 rows and 49 columns as expected
```

Impute missing values with averages within variables "male" and "hispanic"
```{r view missing data}
#look at which columns have NA count greater than 0
which(colSums(is.na(dat)) > 0)
```

```{r impute nas with means of each group}

#select most important columns to impute NAs and assign to "in_names"
in_names  <- c("agepft", "height", "weight", "bmi", "smoke", "asthma", "gasstove", "fev")
out_names <- paste0(in_names, "_avg") #suffix for "out_names" will be _avg

dat <- merge(
  
  #group by variables "hispanic" and "male", then compute the mean of in_names for each group and assign to temporary variable x
  x = dat[,
  setNames(lapply(.SD, mean, na.rm = TRUE), out_names),
  .SDcols = in_names, 
  by  = .(hispanic, male)
  ],
  
  #merge temporary variable x back with main data table to add new columns containing averages
  y = dat,
  by.x = c("hispanic", "male"),
  by.y = c("hispanic", "male")
)

#check results
head(dat[, hispanic:educ_parent]) %>%
  knitr::kable()

```
The imputed average columns have been successfully added to the data table.
\

Next, impute NA values with averages from *_avg columns
```{r impute NAs with averages from _avg columns}
dat[,
     c("agepft", "height", "weight", "bmi", "smoke", "asthma", "gasstove", "fev") :=
     .(
       fifelse(is.na(agepft), agepft_avg, agepft),
       fifelse(is.na(height), height_avg, height),
       fifelse(is.na(weight), weight_avg, weight),
       fifelse(is.na(bmi), bmi_avg, bmi),
       fifelse(is.na(smoke), smoke_avg, smoke),
       fifelse(is.na(asthma), asthma_avg, asthma),
       fifelse(is.na(gasstove), gasstove_avg, gasstove),
       fifelse(is.na(fev), fev_avg, fev)
     )
     ]

#check that there are no more NAs in the important columns
sum(is.na(dat[, ..in_names]))

```

###### 2. Create a new categorical variable named “obesity_level” using the BMI measurement (underweight BMI<14; normal BMI 14-22; overweight BMI 22-24; obese BMI>24). To make sure the variable is rightly coded, create a summary table that contains the minimum BMI, maximum BMI, and the total number of observations per category.


```{r create obesity_level category}
#removing the averages columns created for previous step
dat <- dat %>% select(!ends_with("_avg")) %>% collect()
dat <- as.data.table(dat)

dat[, 
    obesity_level := fifelse(bmi < 14, "underweight", 
                             fifelse(bmi >= 14 & bmi < 22, "normal", 
                                     fifelse(bmi >= 22 & bmi <= 24, "overweight", "obese")))]
```

Create summary table to verify new column.
```{r summary table of new column}
dat[, .(min_bmi = min(bmi), max_bmi = max(bmi)), by = obesity_level][order(min_bmi)] %>%
  knitr::kable()
```
The bmi variables have been categorized as intended.


###### 3. Create another categorical variable named “smoke_gas_exposure” that summarizes “Second Hand Smoke” and “Gas Stove.” The variable should have four categories in total.

As some values will have been imputed with averages, I will view the variables
```{r consider imputed values}
dat[(smoke %between% c(0.001,0.999)) & (gasstove %between% c(0.001,0.999)), .(smoke, gasstove)]
```
It appears that the imputed average for smoke exposure is closer to 0, and the imputed average for gas exposure is closer to 1. When creating the smoke and gas exposure category, those that have smoke exposure = 1 will qualify. As for the gas exposure, those that have gasstove != 0 will qualify. Essentially, I will be rounding the smoke exposure values up to 1 and the gasstove exposure values down to 0.

```{r create smoke/gas category}
dat[,
    smoke_gas_exposure := fifelse(smoke == 1 & gasstove != 0, "both",
                                  fifelse(smoke != 1 & gasstove != 0, "gas",
                                          fifelse(smoke == 1 & gasstove == 0, "smoke","neither")))]
```


###### 4. Create four summary tables showing the average (or proportion, if binary) and sd of “Forced expiratory volume in 1 second (ml)” and asthma indicator by town, sex, obesity level, and “smoke_gas_exposure.”

```{r four summary tables}

#fev and asthma by town name
dat[, .(avg_fev = mean(fev), std_fev = sd(fev), prop_asthma = mean(asthma), asthma_std = sd(asthma)), by = townname] %>%
  knitr::kable()

#fev and asthma by sex
dat[, .(avg_fev = mean(fev), std_fev = sd(fev), prop_asthma = mean(asthma), asthma_std = sd(asthma)), by = male] %>%
  knitr::kable()

#fev and asthma by obesity level
dat[, .(avg_fev = mean(fev), std_fev = sd(fev), prop_asthma = mean(asthma), asthma_std = sd(asthma)), by = obesity_level] %>%
  knitr::kable()

#fev and asthma by smoke and gas exposure
dat[, .(avg_fev = mean(fev), std_fev = sd(fev), prop_asthma = mean(asthma), asthma_std = sd(asthma)), by = smoke_gas_exposure] %>%
  knitr::kable()
```


### Looking at the Data (EDA)
The primary questions of interest are: \
1. What is the association between BMI and FEV (forced expiratory volume)? \
2. What is the association between smoke and gas exposure and FEV? \
3. What is the association between PM2.5 exposure and FEV? \

Follow the EDA checklist from week 3 and the previous assignment. Be sure to focus on the key variables. Visualization Create the following figures and interpret them. Be sure to include easily understandable axes, titles, and legends. \

Check dimensions, headers, footers, and variable types.

```{r check dimensions of data frame}
dim(dat)
```
Dimensions look correct, the number of observations remained the same as the earlier check, and there are two additional columns (for the two categorical variables that were created)

```{r check header and footer}
head(dat) %>%
  knitr::kable()

tail(dat) %>%
  knitr::kable()
```
```{r check variable types}
str(dat)
```
The variables appear to be stored in the correct data type.

\
Take closer look at key variables and calculate initial summary statistics.

The key variables are fev, smoke and gas exposure, pm 2.5, and bmi.

According to this article from NCBI: \ https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2643211/ \
the forced expiratory volume of an individual varies by age and gender. First, I need to know what the age range is for individuals in this data set.
```{r check age range}
summary(dat[, agepft])
```
After verifying on https://github.com/USCbiostats/data-science-data/blob/master/01_chs/chs_codebook.docx that the age numbers are in fact in years, I looked back to the NCBI article to find what range of fev I should expect to see for individuals at ages 8 to 13. According to the plot in Figure 1, the median fev values found for individuals ages 8 to 13 ranged from below 1 liter to about 3 liters for females and 1 liter to 3.5 liters in males.

```{r check fev}
#according to the document describing the data, the fev is recorded in ml.

dat[, .(min = min(fev), median = median(fev), max = max(fev)), by = male]

```
The stats for fev in males and females falls within the range that I would expect them to.



######  1. Facet plot showing scatterplots with regression lines of BMI vs FEV by “townname”.\

```{r facet plot of bmi vs fev}
ggplot(dat, aes(bmi, fev, color = townname)) +
  geom_jitter(size = 0.5) +
  geom_smooth(method = "lm", formula = y~x, size = 0.5, se = FALSE, color = "red") +
  facet_wrap(~ townname, nrow = 3)
```
\
The plots above show that forced expiratory volume and body mass index are positively correlated in each of the towns in the data set. Additionally, towns such as Alpine and Riverside have a relatively low positive slope, whereas towns such as San Dimas, Lake Elsinore, and Upland have steeper positive slopes. This suggests that the positive correlation between body mass index and forced expiratory volume varies slightly by region.

######   2. Stacked histograms of FEV by BMI category and FEV by smoke/gas exposure. Use different color schemes than the ggplot default.

```{r histograms of fev by bmi}

ggplot(data = dat, mapping = aes(x = fev)) +
  geom_histogram(mapping = aes(fill = factor(obesity_level)), bins = 70, position = "dodge", binwidth = 50) +
  scale_fill_manual(values = c("darkseagreen2", "dodgerblue", "firebrick2", "gold1"))
```

```{r histogram of fev by smoke/gas}
ggplot(data = dat, mapping = aes(x = fev)) +
  geom_histogram(mapping = aes(fill = factor(smoke_gas_exposure)), bins = 70, position = "dodge", binwidth = 50) +
  scale_fill_manual(values = c("dodgerblue4", "lightpink", "cadetblue1", "yellow1"))

```
\
Of the two histograms above, the one on top reveals that most people in the data set fall under the normal obesity level category, and that the average forced expiratory volume (fev) for people in this category lies at about 2000 with a normal distribution. As for people in the underweight category, the average fev is lower than that of those in the normal category, at about 1750. For people in both the overweight and obese categories, the average fev lies about at 2300, which is higher than the average for the people in the normal category.\
The histogram on the bottom reveals that most people in the data set are classified as either having exposure to both gas and smoke, or gas only. Additionally, all of the categories have an average fev of about 2000, with a standard deviation of about 300. 

######  3. Barchart of BMI by smoke/gas exposure. \
```{r bmi vs smoke/gas barchart}
ggplot(data = dat, mapping = aes(x = obesity_level, fill = factor(smoke_gas_exposure))) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("dodgerblue4", "lightpink", "cadetblue1", "darkorange1"))
```
\
From this bar chart of bmi category counts by obesity level, it is evident that the proportions of people that are exposed to gas, smoke, both, or neither are mostly consistent across the bmi categories. The highest proportion of people in each category have some exposure to gas, and the smallest proportion of people have some exposure to smoke. For the other types of exposure (both or neither), the proportion of people in those categories sits at about one third of that of the proportion of people exposed to gas only.

######  4. Statistical summary graphs of FEV by BMI and FEV by smoke/gas exposure category.\
```{r plot stat summary of fev by bmi}
ggplot(data = dat, mapping = aes(x = fev, y = bmi, col = obesity_level)) +
  geom_jitter() +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE, size = 0.75) +
  geom_smooth(mapping = aes(x = fev, y= bmi), col = "black", method = "lm", formula = y ~ x, se = FALSE, size = 0.75, linetype = "twodash")
```
\
The above plot reveals that there is a positive correlation between increase in bmi and increase in forced expiratory volume, as shown in the dashed black regression line drawn through the points. Furthermore, within each of the obesity level categories, there seems to be the steepest positive correlation slope for the relationship between bmi and fev for those in the normal obesity level category, followed by obese, then underweight, and finally overweight.

```{r plot stat summary of fev by smoke/gas exposure}

ggplot(data = dat, mapping = aes(y = fev, x = smoke_gas_exposure, color = smoke_gas_exposure)) +
  stat_summary(fun.min = min, fun.max = max, fun = median)
```
\
The stat summary above reveals that when looking at fev by smoke and gas exposure, there the median fev for the groups all lie at 2000. The groups having no exposure to smoke or gas, as well as the group exposed to smoke only, have medians slightly higher than the other two groups. Additionally, the range of observations for the groups exposed to neither smoke nor gas, and gas only, are greater than for the other two groups.

######  5. A leaflet map showing the concentrations of PM2.5 mass in each of the CHS communities.\
```{r plot leaflet map, warning=FALSE}
qpal <- colorQuantile("YlOrRd", dat[, pm25_mass], n = 12)

qpal_colors <- unique(qpal(sort(dat[, pm25_mass])))
qpal_labs <- unique(sort(dat[, pm25_mass]))

leaflet(dat) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircles(lat = ~lat, lng = ~lon, color = ~qpal(pm25_mass), radius = 400, fillOpacity = 1) %>%
  addLegend(colors = qpal_colors, labels = qpal_labs, title = "conc. of PM 2.5 mass", position = "topright", opacity = 0.5)
```
\
The leaflet plot above reveals that higher concentration levels of pm 2.5 are associated with large, urban population centers, in this case around the greater Los Angeles area. Only in areas such as Santa Barbara or San Luis Obispo that are further from large population centers and closer to the coastline are low concentration levels of pm 2.5 of below 8 observed.

######  6. Choose a visualization to examine whether PM2.5 mass is associated with FEV.\
```{r pm25 mass and fev association grouped boxplots}
ggplot(data = dat, mapping = aes(y = fev, x = pm25_mass, group = factor(pm25_mass))) +
  geom_boxplot(width = 0.5) +
  geom_smooth(mapping = aes(group=1), col = "red", formula = y ~ x, method = "lm", alpha = 0.01, linetype = "twodash")
```
\
The above boxplots show that there is a small negative correlation between forced expiratory volume and concentration of pm 2.5 mass, with a slight decrease in fev with increasing concentration of pm 2.5 mass.
