---
title: "Assignment 02: Data Viz and Wrangling"
author: "Flemming Wu"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true
---

For this assignment, we will be analyzing data from USC’s Children’s Health Study. The learning objectives are to conduct data wrangling and visualize the data with key questions in mind.

### Data Wrangling
You will need to download two datasets from https://github.com/USCbiostats/data-science-data. The individual and regional CHS datasets in 01_chs. The individual data includes personal and health characteristics of children in 12 communities across Southern California. The regional data include air quality measurements at the community level. Once downloaded, you can merge these datasets using the location variable. Once combined, you will need to do the following:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
setwd("~/Desktop/PM566/GitHub/02-assignment")
```

Load libraries
```{r load libraries, message = FALSE}
library(data.table)
library(tidyverse)
library(dtplyr)
library(leaflet)
```

Download data
```{r download data}
if(!file.exists("individual.csv")){
  download.file(url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/01_chs/chs_individual.csv",
               destfile = "individual.csv",
               method = "libcurl",
               timeout = 60)
}

if(!file.exists("regional.csv")){
  download.file(url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/01_chs/chs_regional.csv",
               destfile = "regional.csv",
               method = "libcurl",
               timeout = 60)
}

#Read data tables
individual <- fread("individual.csv")
regional <- fread("regional.csv")

```

######  1. After merging the data, make sure you don’t have any duplicates by counting the number of rows. Make sure it matches.\
In the case of missing values, impute data using the average within the variables “male” and “hispanic.” If you are interested (and feel adventurous) in the theme of Data Imputation, take a look at this paper on “Multiple Imputation” using the Amelia R package here.\
  
Merge data tables and ensure dimensions are correct.
```{r merging data}
#Check rows
dim(individual) #merged data table should contain 1200 rows
dim(regional) #merged data table should have 23 + 26 columns

dat <- merge(x = individual, y = regional, by.x = "townname", by.y = "townname")

#Check dimensions
dim(dat) #1200 rows and 49 columns as expected
```

Impute missing values with averages within variables "male" and "hispanic"
```{r view missing data}
#get columns that have na count greater than 0, and pass them into a vector
which(colSums(is.na(dat)) > 0)
```

```{r impute nas with means of each group}

#select most important columns to impute NAs and assign to "in_names"
in_names  <- c("agepft", "height", "weight", "bmi", "smoke", "asthma", "gasstove", "fev")
out_names <- paste0(in_names, "_avg")
dat <- merge(
  
  #group by variables hispanic and male, then compute the mean of in_names for each group and assign to temporary variable x
  x = dat[,
  setNames(lapply(.SD, mean, na.rm = TRUE), out_names),
  .SDcols = in_names, 
  by  = .(hispanic, male)
  ],
  
  #merge temporary variable x back with main data table to add new columns containing averages
  y = dat,
  by.x = c("hispanic", "male"),
  by.y = c("hispanic", "male")
)

head(dat, n = 3)

#impute NAs with averages from _avg columns
dat[,
     c("agepft", "height", "weight", "bmi", "smoke", "asthma", "gasstove", "fev") :=
     .(
       fifelse(is.na(agepft), agepft_avg, agepft),
       fifelse(is.na(height), height_avg, height),
       fifelse(is.na(weight), weight_avg, weight),
       fifelse(is.na(bmi), bmi_avg, bmi),
       fifelse(is.na(smoke), smoke_avg, smoke),
       fifelse(is.na(asthma), asthma_avg, asthma),
       fifelse(is.na(gasstove), gasstove_avg, gasstove),
       fifelse(is.na(fev), fev_avg, fev)
     )
     ]

#check that there are no more NAs in the important columns
sum(is.na(dat[, ..in_names]))

```

###### 2. Create a new categorical variable named “obesity_level” using the BMI measurement (underweight BMI<14; normal BMI 14-22; overweight BMI 22-24; obese BMI>24). To make sure the variable is rightly coded, create a summary table that contains the minimum BMI, maximum BMI, and the total number of observations per category.


```{r create obesity_level category}
#removing the averages columns created for previous step
dat <- dat %>% select(!ends_with("_avg")) %>% collect()
dat <- as.data.table(dat)

dat[, 
    obesity_level := fifelse(bmi < 14, "underweight", 
                             fifelse(bmi >= 14 & bmi < 22, "normal", 
                                     fifelse(bmi >= 22 & bmi <= 24, "overweight", "obese")))]
```

Create summary table to verify new column.
```{r summary table of new column}
dat[, .(min_bmi = min(bmi), max_bmi = max(bmi)), by = obesity_level][order(min_bmi)]
```



###### 3. Create another categorical variable named “smoke_gas_exposure” that summarizes “Second Hand Smoke” and “Gas Stove.” The variable should have four categories in total.
```{r consider imputed values}
dat[(smoke %between% c(0.001,0.999)) & (gasstove %between% c(0.001,0.999)), .(smoke, gasstove)]
```
It appears that the imputed average for smoke exposure is closer to 0, and the imputed average for gas exposure is closer to 1. When creating the smoke and gas exposure category, those that have smoke exposure = 1 will qualify. As for the gas exposure, those that have gasstove != 0 will qualify.

```{r create smoke/gas category}
dat[,
    smoke_gas_exposure := fifelse(smoke == 1 & gasstove != 0, "both",
                                  fifelse(smoke != 1 & gasstove != 0, "gas",
                                          fifelse(smoke == 1 & gasstove == 0, "smoke","neither")))]
```


###### 4. Create four summary tables showing the average (or proportion, if binary) and sd of “Forced expiratory volume in 1 second (ml)” and asthma indicator by town, sex, obesity level, and “smoke_gas_exposure.”

```{r four summary tables}

#fev and asthma by town name
dat[, .(avg_fev = mean(fev), std_fev = sd(fev), avg_asthma = mean(asthma), asthma_std = sd(asthma)), by = townname]

#fev and asthma by sex
dat[, .(avg_fev = mean(fev), std_fev = sd(fev), avg_asthma = mean(asthma), asthma_std = sd(asthma)), by = male]

#fev and asthma by obesity level
dat[, .(avg_fev = mean(fev), std_fev = sd(fev), avg_asthma = mean(asthma), asthma_std = sd(asthma)), by = obesity_level]

#fev and asthma by smoke and gas exposure
dat[, .(avg_fev = mean(fev), std_fev = sd(fev), avg_asthma = mean(asthma), asthma_std = sd(asthma)), by = smoke_gas_exposure]
```


### Looking at the Data (EDA)
The primary questions of interest are: \
1. What is the association between BMI and FEV (forced expiratory volume)? \
2. What is the association between smoke and gas exposure and FEV? \
3. What is the association between PM2.5 exposure and FEV? \

Follow the EDA checklist from week 3 and the previous assignment. Be sure to focus on the key variables. Visualization Create the following figures and interpret them. Be sure to include easily understandable axes, titles, and legends. \

######  1. Facet plot showing scatterplots with regression lines of BMI vs FEV by “townname”.\

```{r facet plot of bmi vs fev}
ggplot(dat, aes(bmi, fev, color = townname)) +
  geom_jitter(size = 0.5) +
  geom_smooth(method = "lm", formula = y~x, size = 0.5, se = FALSE, color = "red") +
  facet_wrap(~ townname, nrow = 3)
```
\
The plots above show that forced expiratory volume and body mass index are positively correlated in each of the towns in the data set. Additionally, towns such as Alpine and Riverside have a relatively low positive slope, whereas towns such as San Dimas, Lake Elsinore, and Upland much steeper positive slopes. This suggests that the positive correlation between bmi and fev varies by region.

######   2. Stacked histograms of FEV by BMI category and FEV by smoke/gas exposure. Use different color schemes than the ggplot default.

```{r histograms of fev by bmi and smoke/gas}
require(gridExtra)

p1 <- ggplot(data = dat, mapping = aes(x = fev)) +
  geom_histogram(mapping = aes(fill = factor(obesity_level)), bins = 70) +
  scale_fill_manual(values = c("darkseagreen2", "lemonchiffon2", "lightsalmon1", "gold1"))

p2 <- ggplot(data = dat, mapping = aes(x = fev)) +
  geom_histogram(mapping = aes(fill = factor(smoke_gas_exposure)), bins = 70) +
  scale_fill_manual(values = c("dodgerblue4", "lightpink", "cadetblue1", "darkorange1"))

grid.arrange(p1, p2, ncol = 1)
```
\
Of the two histograms above, the one on top reveals that most people in the data set fall under the normal obesity level category, and that the average forced expiratory volume (fev) for people in this category lies at about 2000 with a normal distribution. As for people in the underweight category, the average fev is lower than that of those in the normal category, at about 1750. For people in both the overweight and obese categories, the average fev lies about at 2300, which is higher than the average for the people in the normal category.\
The histogram on the bottom reveals that most people in the data set are classified as either having exposure to both gas and smoke, or gas only. Additionally, most of the people in each category have an average fev of about 2000, with a standard deviation of about 300. 

######  3. Barchart of BMI by smoke/gas exposure. \
```{r bmi vs smoke/gas barchart}
ggplot(data = dat, mapping = aes(x = obesity_level, fill = factor(smoke_gas_exposure))) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("dodgerblue4", "lightpink", "cadetblue1", "darkorange1"))
```
\
From this bar chart of bmi category counts by obesity level, it is evident that the proportions of people that are exposed to gas, smoke, both, or neither are mostly consistent across the bmi categories. The highest proportion of people in each category have some exposure to gas, and the smallest proportion of people have some exposure to smoke. For the other types of exposure (both or neither), the proportion of people in those categories sits at about one third of that of the proportion of people exposed to gas only.

######  4. Statistical summary graphs of FEV by BMI and FEV by smoke/gas exposure category.\
```{r plot stat summary of fev by bmi}
ggplot(data = dat, mapping = aes(x = fev, y = bmi, col = obesity_level)) +
  geom_jitter() +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE)
```
```{r plot stat summary of fev by smoke/gas exposure}

ggplot(data = dat, mapping = aes(y = fev, x = smoke_gas_exposure, color = smoke_gas_exposure)) +
  geom_violin(trim=FALSE, fill = NA) +
  stat_summary(fun.min = min, fun.max = max, fun = median)
```

######  5. A leaflet map showing the concentrations of PM2.5 mass in each of the CHS communities.\
```{r plot leaflet map, warning=FALSE}
qpal <- colorQuantile("YlOrRd", dat[, pm25_mass], n = 12)

qpal_colors <- unique(qpal(sort(dat[, pm25_mass])))
qpal_labs <- unique(sort(dat[, pm25_mass]))

leaflet(dat) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircles(lat = ~lat, lng = ~lon, color = ~qpal(pm25_mass), radius = 200, fillOpacity = 1) %>%
  addLegend(colors = qpal_colors, labels = qpal_labs, title = "conc. of PM 2.5 mass")
```


######  6. Choose a visualization to examine whether PM2.5 mass is associated with FEV.\
```{r pm25 mass and fev association visualization}
library(ggbeeswarm)

ggplot(data = dat, mapping = aes(x = pm25_mass, y = fev, col = pm25_mass)) +
  geom_beeswarm() +
  geom_smooth(formula = y ~ x, method = "lm", col = "black")
```

