---
title: "Lab 03 - Exploratory Data Analysis"
author: "Flemming Wu"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Steps

### 1. Read in the data
```{r, message = FALSE}
if(!file.exists("met_all.gz")){
  download.file("https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz", "met_all.gz", method="libcurl", timeout = 60)
}
met <- data.table::fread("met_all.gz")
```

### 2. Check the dimensions, headers, footers. How many columns, rows are there?
```{r, check dimensions}
dim(met)
```
There are 2,377,343 million rows/observations, and 30 columns/variables

```{r, check head}
head(met)
```
```{r, check tail}
tail(met)
```

### 3. Take a look at the variables.

```{r}
str(met)
```

### 4. Take a closer look at the key variables.

```{r}
table(met$year)
```

```{r}
table(met$day)
```

```{r}
table(met$hour)
```

```{r}
summary(met$temp)
```
Improbable low minimum temperature value.

```{r}
summary(met$elev)
```
The 9999 variable for the max probably indicates missing value(s).

```{r}
summary(met$wind.sp)
```
Lots of missing data.

##### Replace 9999 elevation data with appropriate representation.
```{r}
met[met$elev==9999.0] <- NA
summary(met$elev)
```
The elevation at the highest weather station is 4113.

##### Remove -40 degree temperature observations
```{r}
met <- met[met$temp > -40]
met2 <- met[order(temp)]
head(met2)
```
The -17.2 temp observation also seems sus.


### 5. Check the data against an external data source.

##### Remove the -17.2 observation.
```{r}
met <- met[met$temp>-15]
met2 <- met[order(temp)]
head(met2)
```
```{r}
summary(met$temp)
```
Minimum temp of -3 degrees Celsius (~27 degrees Fahrenheit) is reasonable.

### 6. Calculate summary statistics

##### Examine windspeed and temperature for station with maximum elevation.
##### Select the weather station with maximum elevation; look at the correlation between temperature and wind speed; look at the correlation between temperature and wind speed with hour and day of the month.

```{r}
elev <- met[met$elev==max(elev)]
summary(elev)
```
```{r}
cor(elev$temp, elev$wind.sp, use="complete")
```
```{r}
cor(elev$temp, elev$hour, use="complete")
```
```{r}
cor(elev$wind.sp, elev$day, use="complete")
```
```{r}
cor(elev$wind.sp, elev$hour, use="complete")
```
```{r}
cor(elev$temp, elev$day, use="complete")
```

### 7. Exploratory graphs

```{r}
hist(met$elev, breaks=100)
```
Skewed right, with few elevation observations above 2000 meters.

```{r}
hist(met$temp)
```
```{r}
hist(met$wind.sp)
```
This data is also skewed, may help to transform it for later analyses.

```{r, Loading in leaflet, message = FALSE}
if(!require(leaflet)){
  install.packages("leaflet")
}

library(leaflet)
```

```{r}
leaflet(elev) %>%
  addProviderTiles("OpenStreetMap") %>%
  addCircles(lat=~lat, lng=~lon, opacity=1, fillOpacity=1, radius=100)
```

##### Look at time series of temperature and wind speed. Create date-time variable for x-axis.

```{r, message = FALSE}
if(!require(lubridate)){
  install.packages("lubridate")
}
library(lubridate)
```

```{r, create new date column}
elev$date <- with(elev, ymd_h(paste(year, month, day, hour, sep= ' ')))
summary(elev$date)
```
```{r}
elev <- elev[order(date)]
head(elev)
```
```{r, plotting temperature vs date}
plot(elev$date, elev$temp, type="l")
```
The temperature seems to fluctuate between 4 and 12.

```{r, plotting wind speed vs date}
plot(elev$date, elev$wind.sp, type="l")
```

