---
title: "Assignment 03: Web Scraping and Text Mining"
author: "Flemming Wu"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
setwd("~/Desktop/PM566/GitHub/03-assignment")
rm(list=ls())
```

```{r load libs, message=FALSE}
library(httr)
library(rvest)
library(stringr)
library(xml2)
library(tidyverse)
library(tidytext)
library(forcats)
require(gridExtra)
```


### APIs

* Using the NCBI API, look for papers that show up under the term “sars-cov-2 trial vaccine.” Look for the data in the pubmed database, and then retrieve the details of the paper as shown in lab 7. How many papers were you able to find?

* Using the list of pubmed ids you retrieved, download each papers’ details using the query parameter rettype = abstract. If you get more than 250 ids, just keep the first 250.

* As we did in lab 7. Create a dataset containing the following:

  1. Pubmed ID number,
  
  2. Title of the paper,
  
  3. Name of the journal where it was published,
  
  4. Publication date, and
  
  5. Abstract of the paper (if any).
  

```{r find number of papers}
# using rvest
count1 <- rvest::read_html("https://pubmed.ncbi.nlm.nih.gov/?term=sars-cov-2+trial+vaccine") %>%
  html_node(xpath = "/html/body/main/div[9]/div[2]/div[2]/div[1]/div[1]") %>%
  as.character() %>%
  str_extract_all("([:digit:]{1,3},)+?[:digit:]{3}") 

paste0("I have found ", count1[[1]], " papers that show up under the term: sars-cov-2 trial vaccine using rvest")


# using xml2
website <- xml2::read_html("https://pubmed.ncbi.nlm.nih.gov/?term=sars-cov-2+trial+vaccine")

count2 <- xml2::xml_find_first(website, xpath = "/html/body/main/div[9]/div[2]/div[2]/div[1]/div[1]") %>%
  as.character() %>%
  str_extract("([:digit:]{1,3},)+?[:digit:]{3}")


paste0("I have found ", count2, " papers that show up under the term: sars-cov-2 trial vaccine using xml2")
```



```{r query 250 articles from pubmed}
# get list of ID numbers
query_ids <- GET(
  url = "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi",
  query = list(
    term = "sars-cov-2 trial vaccine",
    retmax = 250) 
)

# Extracting the content of the response of GET
ids <- httr::content(query_ids) %>% as.character()

ids <- stringr::str_extract_all(ids, "<Id>[:digit:]*</Id>")[[1]] %>% stringr::str_remove_all("</?Id>")

publications <- GET(
  url   = "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi", # the link now will be using efetch, instead of esearch
  query = list(
    db = "pubmed",
    id = str_c(ids, collapse = ","),
    rettype = "abstract"
    )
)

# Turning the output into character vector
publications <- httr::content(publications) %>% xml2::xml_children()
publications_txt <- as.character(publications)

```



```{r get required informatio from publications text}

# article titles
titles <- str_extract(publications_txt, "<ArticleTitle>.*</ArticleTitle>") %>% str_remove_all("</?[:alpha:]*>")

# names of journals
journals <- str_extract(publications_txt, "<ISOAbbreviation>.*</ISOAbbreviation>") %>% str_remove_all("</?[:alpha:]+>")


# publication dates
dates <- str_remove_all(publications_txt, "\\n") %>% 
  str_extract("<PubDate>.*</PubDate>") %>% 
  str_squish() %>% 
  str_remove_all("</?[:alpha:]+>") %>% 
  str_trim(side = "both")


# abstract texts
abstracts <- str_remove_all(publications_txt, "\\n") %>% 
  str_extract("<Abstract>.{1,}?</Abstract>") %>%
  str_remove_all("(</?AbstractText Label.{1,}?>|</?Abstract>|</?AbstractText>)") %>%
  str_trim(side = "both")

```





```{r create data frame}
df <- data.frame(
  PubMed_Id = ids,
  title = titles,
  journal = journals,
  publication_date = dates,
  abstract = abstracts,
  check.rows = FALSE
)
```

```{r view data frame}
head(df, n = 3) %>%
  knitr::kable()
```



### Text Mining

A new dataset has been added to the data science data repository (https://github.com/USCbiostats/data-science-data/tree/master/03_pubmed)[https://github.com/USCbiostats/data-science-data/tree/master/03_pubmed]. The dataset contains 3241 abstracts from articles across 5 search terms. Your job is to analyse these abstracts to find interesting insights.

1. Tokenize the abstracts and count the number of each token. Do you see anything interesting? Does removing stop words change what tokens appear as the most frequent? What are the 5 most common tokens for each search term after removing stopwords?

```{r read in data from GitHub}
pm <- read_csv("https://raw.githubusercontent.com/USCbiostats/data-science-data/master/03_pubmed/pubmed.csv", col_types = c("c", "c"))
```


```{r tokenize abstracts, message=FALSE}
pm %>%
  unnest_tokens(token, abstract) %>%
  count(token, sort = TRUE) %>%
  top_n(n = 20) %>%
  ggplot(aes(x = n, y = fct_reorder(token, n))) +
  geom_col() +
  labs(y = "Token", title = "Top 20 Most Frequent Words in PubMed Data Set")
```

There is nothing particularly interesting about the most frequent words in the abstract texts, as most of them contain stop words. Some of the non-stop words include "patients", "cancer", and "covid".

```{r removing stop words and numbers, message=FALSE}
pm %>%
  unnest_tokens(token, abstract) %>%
  anti_join(stop_words, by = c("token" = "word")) %>%
  filter( !grepl(pattern = "^[0-9]+$", x = token)) %>%
  count(token, sort = TRUE) %>%
  top_n(n = 30) %>%
  ggplot(aes(x = n, y = fct_reorder(token, n))) +
  geom_col() +
  labs(y = "Token", title = "Top 20 Most Frequent Words in PubMed Data Set (Stop Words and Numbers Removed)") + 
  theme(plot.title = element_text(size = 10))
```

After removing stop words and digits, it is evident that the text contained in the data set are about health related topics, with the most frequent words including: "covid", "patients", "disease", and "eclampsia".

```{r top 5 tokens for each search term after removing stop words, warning=FALSE, fig.width=12, fig.height=12, fig.align='center'}
top_5 <- pm %>%
  unnest_tokens(token, abstract) %>%
  anti_join(stop_words, by = c("token" = "word")) %>%
  filter(!grepl(pattern = "^[0-9]+$", x = token)) %>%
  group_by(term) %>%
  count(token, sort = TRUE) %>%
  top_n(5) %>%
  arrange(term)

p1 <- ggplot(top_5 %>% filter(term == "covid"), aes(n, fct_reorder(token, n))) +
  geom_col() +
  labs(y = "token", title = "Term: covid")

p2 <- ggplot(top_5 %>% filter(term == "cystic fibrosis"), aes(n, fct_reorder(token, n))) +
  geom_col() +
  labs(y = "token", title = "Term: cystic fibrosis")
             
p3 <- ggplot(top_5 %>% filter(term == "meningitis"), aes(n, fct_reorder(token, n))) +
  geom_col() +
  labs(y = "token", title = "Term: meningitis")

p4 <- ggplot(top_5 %>% filter(term == "preeclampsia"), aes(n, fct_reorder(token, n))) +
  geom_col() +
  labs(y = "token", title = "Term: preeclampsia")

p5 <- ggplot(top_5 %>% filter(term == "prostate cancer"), aes(n, fct_reorder(token, n))) +
  geom_col() +
  labs(y = "token", title = "Term: prostate cancer")
             
grid.arrange(p1, p2, p3, p4, p5, ncol = 2)
```

The top 5 words for each of the search terms contain the search terms themselves, but other than that the word "disease" appears frequently under the terms prostate cancer, cystic fibrosis, and covid. The word "patient" appears frequently under the terms prostate cancer, meningitis, cystic fibrosis, and covid. 


2. Tokenize the abstracts into bigrams. Find the 10 most common bigram and visualize them with ggplot2.
Calculate the TF-IDF value for each word-search term combination. (here you want the search term to be the “document”) 

```{r top 10 bigrams by tf-idf}
bigrams <- pm %>%
  unnest_ngrams(ngram, abstract, n = 2) %>%
  count(ngram, term, sort = TRUE) %>%
  bind_tf_idf(ngram, term, n) %>%
  arrange(desc(tf_idf)) %>%
  head(10)
```


```{r plot top 10 bigrams by term frequency only}
bigrams %>%
  top_n(n, n = 10) %>%
  ggplot(aes(x = n, y = fct_reorder(ngram, n))) +
  geom_col() +
  labs(x = "TF-IDF", y = "Bigram", title = "Top 10 Bigrams Ranked by Term Frequency Only")
```



3. What are the 5 tokens from each search term with the highest TF-IDF value? How are the results different from the answers you got in question 1?
