---
title: "Lab 10 - SQL"
author: "Flemming Wu"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
#options(repos = c(CRAN = "http://cran.rstudio.com"))
rm(list=ls())
```

## Setup

```{r install and load libraries, message=FALSE, warning=FALSE}
if(!require("RSQLite")) { install.packages("RSQLite") }
if(!require("DBI")) { install.packages("DBL") }

library(RSQLite)
library(DBI)
```

```{r create db}
# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
actor <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/actor.csv")
rental <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/rental.csv")
customer <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/customer.csv")
payment <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/payment_p2007_01.csv")

# Copy data.frames to database
dbWriteTable(con, "actor", actor)
dbWriteTable(con, "rental", rental)
dbWriteTable(con, "customer", customer)
dbWriteTable(con, "payment", payment)
```

```{r list tables in R memory}
dbListTables(con)
```
TIP: You can use the following QUERY to see the structure of a table


```{sql connection=con}
PRAGMA table_info(actor)
```

SQL references:

[https://www.w3schools.com/sql/](https://www.w3schools.com/sql/)


### Exercise 1
Retrive the actor ID, first name and last name for all actors using the actor table. Sort by last name and then by first name.
```{sql connection=con}
SELECT actor_id, first_name, last_name
FROM actor
ORDER BY last_name, first_name;
```


## Exercise 2
Retrieve the actor ID, first name, and last name for actors whose last name equals ‘WILLIAMS’ or ‘DAVIS’.
```{sql connection=con}
SELECT actor_id, first_name, last_name
FROM actor
WHERE last_name IN ('WILLIAMS', 'DAVIS')
ORDER BY last_name;
```


## Exercise 3
Write a query against the rental table that returns the IDs of the customers who rented a film on July 5, 2005 (use the rental.rental_date column, and you can use the date() function to ignore the time component). Include a single row for each distinct customer ID.
```{sql connection=con}
-- PRAGMA table_info(rental)

SELECT DISTINCT customer_id
FROM rental
WHERE DATE(rental_date) = '2005-07-05';
```


## Exercise 4

### Exercise 4.1
Construct a query that retrieves all rows from the payment table where the amount is either 1.99, 7.99, 9.99.

```{sql connection=con}
SELECT *
FROM payment
WHERE amount IN (1.99, 7.99, 9.99);
```

### Exercise 4.2
Construct a query that retrieves all rows from the payment table where the amount is greater then 5

```{sql connection=con}
SELECT *
FROM payment
WHERE amount > 5;
```

### Exercise 4.3
Construct a query that retrieves all rows from the payment table where the amount is greater than 5 and less than 8

```{sql connection=con}
SELECT *
FROM payment
WHERE amount BETWEEN 5 AND 8;
```

## Exercise 5
Retrieve all the payment IDs and their amount from the customers whose last name is ‘DAVIS’.

```{sql connection=con}
SELECT c.customer_id, c.last_name, p.payment_id, p.amount
FROM payment p
INNER JOIN customer c
ON p.customer_id = c.customer_id
WHERE c.last_name LIKE 'DAVIS';
```

## Exercise 6

### Exercise 6.1
Use COUNT(*) to count the number of rows in rental
```{sql connection=con}
SELECT COUNT(*) num_rows
FROM rental;
```

### Exercise 6.2
Use COUNT(*) and GROUP BY to count the number of rentals for each customer_id
```{sql connection=con}
SELECT customer_id, COUNT(*) num_rentals
FROM rental
GROUP BY customer_id;
```

### Exercise 6.3
Repeat the previous query and sort by the count in descending order
```{sql connection=con}
SELECT customer_id, COUNT(*) num_rentals
FROM rental
GROUP BY customer_id
ORDER BY num_rentals DESC;
```

### Exercise 6.4
Repeat the previous query but use HAVING to only keep the groups with 40 or more.
```{sql connection=con}
SELECT customer_id, COUNT(*) num_rentals
FROM rental
GROUP BY customer_id
HAVING num_rentals > 40
ORDER BY num_rentals DESC;
```

## Exercise 7
The following query calculates a number of summary statistics for the payment table using MAX, MIN, AVG and SUM
```{sql connection=con}
SELECT MAX(amount) max_payment, 
       MIN(amount) min_payment, 
       AVG(amount) avg_payment, 
       SUM(amount) total_payment,
       COUNT(*) num_payments
FROM payment;
```

### Exercise 7.1
Modify the above query to do those calculations for each customer_id
```{sql connection=con}
SELECT customer_id, 
       MAX(amount) max_payment, 
       MIN(amount) min_payment, 
       AVG(amount) avg_payment, 
       SUM(amount) total_payment,
       COUNT(*) num_payments
FROM payment
GROUP BY customer_id;
```

### Exercise 7.2
Modify the above query to only keep the customer_ids that have more then 5 payments
```{sql connection=con}
SELECT customer_id, 
       MAX(amount) max_payment, 
       MIN(amount) min_payment, 
       AVG(amount) avg_payment, 
       SUM(amount) total_payment,
       COUNT(*) num_payments
FROM payment
GROUP BY customer_id
HAVING num_payments > 5;
```

## Cleanup
Run the following chunk to disconnect from the connection.
```{r disconnect connection to database}
dbDisconnect(con)
```

